Modeling Ionic Polymer-Metal Composites\\ with Space-Time Adaptive Multimesh \emph{hp}-FEM

### 2011-03-21 intro.tex
The final version

Ionic Polymer-Metal Composites (IPMC) have been studied during the past two 
decades for their potential to serve as noiseless mechanoelectrical and electromechanical 
transducers \cites{basu1997membrane,shahinpoor2001smartmat,
nasser2002applied,newbury2003intelligent, wallmersperger2007appliedphysics,
pugal2008appliedphysics,pugal2010polymer}.
The advantages of IPMC over other electroactive polymer actuators
are low voltage bending, high strains ($>1\%$), and an ability to work
in wet environments. A typical IPMC consists of a thin sheet of polymer
(often Nafion or Teflon) which is sandwiched between noble
metal electrodes such as platinum or gold. When fabricated, the polymer 
membrane is saturated with certain solvent and ions such as water and $H^+$.
When a voltage is applied to the electrodes, the counter ions start
migrating due to the imposed electric field. By dragging along the solvent,
the osmotic pressure difference near the electrodes
results in bending of the material (see Fig.~\ref{fig:conceptual}).
\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[scale=0.7]{IPMC_bending}
  \caption{\label{fig:conceptual}Conceptual model of the actuation
 	of IPMC. Initial counter ion distribution (a) and
	the distribution and resulting bending after applying a voltage (b).}
  \end{centering}
%\vspace{-1cm}
\end{figure}
%\newpage
%The derived and implemented model helps to predict the actuation of the
%material. Furthermore, it is expected that the \emph{hp}-FEM implementation
%results in a smaller problem size, thus likely allowing faster calculation
%in both 2D and 3D domains.

@@

The outline of the paper is as follows: Section \ref{sec:moti} shows that 
the solution components $C$ and $\phi$ have very different behavior, which
is the reason why it is difficult to find a common mesh that would be optimal 
for both of them. This explains why we are interested in approximation them
on individual meshes equipped with mutually independent adaptivity 
mechanisms. The PNP model is presented in Section \ref{sec:model} where also its weak 
formulation for the Newton's method is derived. Section \ref{sec:hermes}
presents a brief overview of a novel adaptive multimesh $hp$-FEM 
method~\cite{solin2010monolithic,solin2010adaptive,dubcova2010space,solini2010adaptive}
that is used to solve the 
problem numerically. Numerical results and comparisons are presented in Section 
\ref{sec:results}, and conclusion and outlook are drawn in Section \ref{sec:conc}.

>>>>

Ionic Polymer-Metal Composites (IPMC) have been studied during the past two 
decades for their potential to serve as noiseless mechanoelectrical and electromechanical 
transducers \cites{basu1997membrane,shahinpoor2001smartmat,
nasser2002applied,newbury2003intelligent, wallmersperger2007appliedphysics,
pugal2008appliedphysics,pugal2010polymer}.
The advantages of IPMC over other electroactive polymer actuators
are low voltage bending, high strains ($>1\%$), and an ability to work
in wet environments. A typical IPMC consists of a thin sheet of polymer
(often Nafion or Teflon) which is sandwiched between noble
metal electrodes such as platinum or gold. When fabricated, the polymer 
membrane is saturated with certain solvent and ions such as water and $H^+$.
When a voltage is applied to the electrodes, the counter ions start
migrating due to the imposed electric field. By dragging along the solvent,
the osmotic pressure difference near the electrodes
results in bending of the material (see Fig.~\ref{fig:conceptual}).
\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[scale=0.7]{IPMC_bending}
  \caption{\label{fig:conceptual}Conceptual model of the actuation
 	of IPMC. Initial counter ion distribution (a) and
	the distribution and resulting bending after applying a voltage (b).}
  \end{centering}
%\vspace{-1cm}
\end{figure}
%\newpage

@@

The outline of the paper is as follows: Section \ref{sec:moti} shows that 
the solution components $C$ and $\phi$ have very different behavior, which
is the reason why it is difficult to find a common mesh that would be optimal 
for both of them. This explains why we are interested in approximation them
on individual meshes equipped with mutually independent adaptivity 
mechanisms. The PNP model is presented in Section~\ref{sec:model} where also its weak 
formulation for the Newton's method is derived. Section~\ref{sec:hermes}
presents a brief overview of a novel adaptive multi-mesh $hp$-FEM 
method~\cite{solin2010monolithic,solin2010adaptive,dubcova2010space,solini2010adaptive}
that is used to solve the 
problem numerically. Numerical results and comparisons are presented in Section~\ref{sec:results}, 
and conclusion and outlook are drawn in Section~\ref{sec:conc}.

### 2011-03-21 model.tex
The final version

Performing integration by parts and taking into account the boundary 

@@

\subsection{Jacobian matrix and residual vector for the Newton's method}
To employ the Newton's method for the nonlinear system \eqref{eq:nernst-planck-weak-final},
\eqref{eq:poisson-weak-final}, formulas for the Jacobian matrix and residual vector need to
be derived. Time discretization will be performed using the second-order Crank-Nicolson 
method. The unknown solution components 
$c^{n+1}$ and $\varphi^{n+1}$ at the end of the time step $\delta\tau$ are expressed 
as linear combinations of finite element basis functions $v_k^{c}$ 
and $v_k^{\varphi}$ with unknown coefficients,
\begin{eqnarray}
  c^{n+1} = c(Y^{n+1}) &=& \sum_{k=1}^{N^c} y_k^{c} v_k^{c}, \label{eq:cnotation}\\
  \varphi^{n+1} = \varphi(Y^{n+1}) &=& \sum_{k=1}^{N^{\varphi}} y_k^{\varphi} v_k^{\varphi}\label{eq:phinotation}.
\end{eqnarray}
Here $Y^{n+1}$ is a coefficient vector of length $N^c + N^{\varphi}$ comprising the unknown 
solution coefficients $y_k^{C}$ and $y_k^{\varphi}$ (in this order). We will also be 
using $c^n = c(Y^n)$ and $\varphi^n = \varphi(Y^n)$ for the previous time step solutions.

@@

%\newpage

@@

% Pavel
% The reader might loose at this point the interest in reading further
% since there is no motivation which favros a hp-FEM discretization over traditional
% low order FEM. One sentence regarding computational performance
% or a related comment might be useful...
Automatic adaptivity in higher-order finite element methods (\emph{hp}-FEM) 
is much different from adaptivity in low-order FEM. Firstly, analytical
error estimates capable of guiding adaptive $hp$-FEM do not exist even for 
the simplest linear elliptic equations, not speaking about nonlinear multiphysics 
coupled systems. Secondly,
a higher-order element can be refined in many different ways, as illustrated 
in Fig.~\ref{fig:refinements}.
\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=0.5\columnwidth]{refinements}
  \caption{\label{fig:refinements} Many possible refinement candidates for a fourth-order
  element.}
  \end{centering}
%\vspace{-0.5cm}
\end{figure}

>>>>

After performing integration by parts and taking into account the boundary 

@@

\subsection{Jacobian matrix and residual vector for the Newton's method}
To employ the Newton's method for the nonlinear system \eqref{eq:nernst-planck-weak-final},
\eqref{eq:poisson-weak-final}, formulas for the Jacobian matrix and residual vector need to
be derived. Time discretization will be performed using the second-order Crank-Nicolson 
method. The unknown solution components 
$c^{n+1}$ and $\varphi^{n+1}$ at the end of the time step $\delta\tau$ are expressed 
as linear combinations of finite element basis functions $v_k^{c}$ 
and $v_k^{\varphi}$ with unknown coefficients,
\begin{eqnarray}
  c^{n+1} = c(Y^{n+1}) &=& \sum_{k=1}^{N^c} y_k^{c} v_k^{c}, \label{eq:cnotation}\\
  \varphi^{n+1} = \varphi(Y^{n+1}) &=& \sum_{k=1}^{N^{\varphi}} y_k^{\varphi} v_k^{\varphi}\label{eq:phinotation}.
\end{eqnarray}
Here $Y^{n+1}$ is a coefficient vector of length $N^c + N^{\varphi}$ comprising the unknown 
solution coefficients $y_k^{c}$ and $y_k^{\varphi}$ (in this order). We will also be 
using $c^n = c(Y^n)$ and $\varphi^n = \varphi(Y^n)$ for the previous time step solutions.

@@

@@

Automatic adaptivity in higher-order finite element methods (\emph{hp}-FEM) 
is much different from adaptivity in low-order FEM. Firstly, analytical
error estimates capable of guiding adaptive $hp$-FEM do not exist even for 
the simplest linear elliptic equations, not speaking about nonlinear multiphysics 
coupled systems. Secondly,
a higher-order element can be refined in many different ways, as illustrated 
in Fig.~\ref{fig:refinements}.
\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=0.5\columnwidth]{refinements}
  \caption{\label{fig:refinements} Many possible refinement candidates for a fourth-order
  element.}
  \end{centering}
%\vspace{-0.5cm}
\end{figure}

### 2011-03-21 pugal-cicp2010.tex
Revision1 done
The final version

%\journal{Applied Mathematics and Computation}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsrefs}
%\usepackage{subfig}

@@

%\title{\emph{hp}-Finite Element Model of Poisson and Nernst-Planck\\ System of Equations}
\title{Modeling Ionic Polymer-Metal Composites\\ with Space-Time Adaptive Multimesh \emph{hp}-FEM}

@@

%\author[unrme,tartu]{D.~Pugal}
%\ead{david.pugal@gmail.com}
%
%\author[unrmath,czech]{P.~Solin\corref{cor2}}
%\ead{solin@unr.edu}
%
%\author[unrme]{K.~J.~Kim\corref{cor1}}
%\ead{kwangkim@unr.edu}
%
%\author[tartu]{A.~Aabloo}
%\ead{alvo@ut.ee}
%
%\address[unrme]{Mechanical Engineering Department, University of Nevada, Reno, NV, U.S.A.}
%\address[unrmath]{Department of Mathematics and Statistics,
%University of Nevada, Reno, NV, U.S.A.}
%\address[czech]{Institute of Thermomechanics, Prague, Czech Republic}
%\address[tartu]{Institute of Technology, Tartu University, Estonia}
%
%\cortext[cor1]{Corresponding author}
%\cortext[cor2]{Principal corresponding author}

@@

%\begin{keyword}
%\keywords{Ionic polymer-metal composites, IPMC, Nernst-Planck equation, Poisson equation,
%finite element method, space-time adaptivity, multimesh \emph{hp}-FEM }
%\end{keyword}

%see http://www.ams.org/mathscinet/msc/msc2010.html?t=35Qxx&btn=Current

\ams{35Q84, 35Q05, 35J47}
\keywords{Ionic polymer-metal composites, IPMC, Nernst-Planck equation, Poisson equation,
finite element method, FEM, adaptive multimesh \emph{hp}-FEM }
\maketitle

%\end{frontmatter}

@@

%\begin{thebibliography}{99}
%\bibitem{\ldots}..

%\end{thebibliography}

>>>>

\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsrefs}

@@

\title{Modeling Ionic Polymer-Metal Composites\\ with Space-Time Adaptive Multimesh \emph{hp}-FEM}

@@

@@

\ams{35Q84, 35Q05, 35J47}
\keywords{Ionic polymer-metal composites, IPMC, Nernst-Planck equation, Poisson equation,
finite element method, FEM, adaptive multi-mesh \emph{hp}-FEM }
\maketitle

@@

\newpage
